Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-16 Manuscript/manuscript.tex
Some updates in the results and discussion, not ready yet

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

@@

To quantify the ion binding affinity to a lipid bilayer, we
calculated the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
This quantity was chosen because it does not depend on
%the definition of the interfacial region or
the position of the Gibbs dividing plane between two bulk regions.
%\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
%SAMULI: Here is a new trial.
Here we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating the order parameters
for C-H bonds and the scattering form factors. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model
\todo{More justification for the choosing the OPC3 water model are needed.
It might be good to show the comparison with the scattering data in bulk water in SI.}. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
\todo{The normal TIP3P was tested as well, right?} presented in Supporting Information (SI). 
The ECC-ion models were used for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulation with the Lipid14 model was also ran with the ion model by Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ion model
was taken directly from \cite{lipid14POPC0mMNaClfiles}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments by using the electrometer concept, because 
the order parameters decrease proportionally to the amount of bound positive charge~\cite{seelig87,catte16}. 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate the Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduce
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions,
as seen from the data replotted from Ref.~\citenum{catte16} in Fig.~\ref{fig:delta_ordPar_CaCl}.
The model reproduced the experimentally measured small order parameter changes with NaCl
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ 
concentration were overestimated by the same combination of force field parameters.

@@

%To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
%polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
Our simulations with the ECC-lipid and the ECC-ion models \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
exhibit an improved behaviour of cation binding to a POPC bilayer
showing a good agreement with experiments in the changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations
in Fig.~\ref{fig:delta_ordPar_CaCl}. Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipid model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.
%In conclusion, the results suggest that 
%the ECC-lipid model together with the ECC-ion parameters gives a realistic
%description for the binding affinities of Na$^+$ and Ca$^{2+}$ ions to POPC lipid bilayer. 
Furthermore, the overestimated headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models considered in the previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

@@

To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. 
Calcium cations at a molar concentration of $350\,\mathrm{mM}$ provide
$\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$ in the ECC-lipid model, which is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to previous works employing classical MD simulations,
which generally agree together on the existence of an increased sodium density at the membrane-water interface.
\todo{JOE: Add references to NaCl simulation papers that show peaks.
Add a simple note that there is NBFIX trick in Charmm36 (?).} 
This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
which show the density profile of ions and water along the membrane normal. 
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}

@@

We can estimate relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}} and \ce{Na+} cations 
using the number of contacts per lipid between cations and POPC oxygen atoms, 
$\Sigma ^\mathrm{Na^{2+}} _\mathrm{O}$,
in which a contact between the two atoms is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ 
(robustness of this definition discussed in reference~\citenum{javanainen17}).

For sodium at $1000\,mM$ concentration 
the average number of contacts per lipid with any oxygen atom in POPC is 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
this number decreases to 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are counted 
it is even lower, 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
This means that interaction of sodium and POPC 
has balanced contributions of 55\% from only phosphate oxygens 
(i.e. no contacts with carbonyl oxyens),
and from only carbonyl oxygens with 20\% of configurations 
leaving 25\% for mixed phosphate and carbonyl oxygen configurations.  
This is in line with the calculated negative value of 
relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
suggesting that the populations of bound sodium cations 
naturally decrease with the decreasing density of water.

@@

>>>>

The concept can be used to compare the ion binding affinity to lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

@@

To quantify the ion binding affinity to a lipid bilayer, we
calculated the relative surface excess of ions with respect to water,~$\Gamma_i^w$~\cite{chattorajBOOK}.
This quantity was chosen because it does not depend on
%the definition of the interfacial region or
the position of the Gibbs dividing plane between two bulk regions.
%\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
%SAMULI: Here is a new trial.
Here we assume that the interface locates between the hydrophobic interior
of a lipid bilayer and the bulk water region far from the membrane.
The bulk concentration of ions and water is zero inside the bilayer.
The concentrations in bulk water region can be calculated from 
the farthest point from both lipid leaflets in the water phase.
The region between these boundaries contains all the ions and water
molecules in the simulation box.
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the aqueous phase;
and $A_b$ is the area of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface at both leaflets.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating the order parameters
for C-H bonds and the scattering form factors. The former validates the structures sampled by the individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model
\todo{More justification for the choosing the OPC3 water model are needed.
It might be good to show the comparison with the scattering data in bulk water in SI.}. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
\todo{The normal TIP3P was tested as well, right?} presented in Supporting Information (SI). 
The ECC-ion models were used for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulation with the Lipid14 model was also ran with the ion model by Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ion model
was taken directly from \cite{lipid14POPC0mMNaClfiles}. \todo{Which water model was used in these simulations?} 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments by using the electrometer concept, because 
the order parameters decrease proportionally to the amount of bound positive charge~\cite{seelig87,catte16}. 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate the Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduce
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions,
as seen from the data replotted from Ref.~\citenum{catte16} in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}.
The model reproduced the experimentally measured small order parameter changes with NaCl
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ 
concentration were overestimated by the same combination of force field parameters.

@@

%To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
%polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
The results from the simulations with the ECC-lipid and the ECC-ion models \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
exhibit an improved behaviour of cation binding to a POPC bilayer in Fig.~\ref{fig:delta_ordPar_CaCl},
showing a good agreement with experiments in the changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations.
Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipid model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.
%In conclusion, the results suggest that 
%the ECC-lipid model together with the ECC-ion parameters gives a realistic
%description for the binding affinities of Na$^+$ and Ca$^{2+}$ ions to POPC lipid bilayer. 
Furthermore, the overestimated lipid headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models considered in the previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

@@

\begin{table}
  \caption{The relative surface excess and the average number of contacts of Calcium with different
    oxygens in POPC with different simulation models \label{tab:binding} }
  \begin{tabular}{l|c c c}
    model                 & $\Gamma_{Ca}^w$ (nm$^{-2}$)   & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.07 \pm 0.01 $            &  0.133   &    0.033    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $            &  -       &     -       \\
    Lipid14/Dang          &  $ 0.3 \pm 0.03  $           &  -        &    -       \\
  \end{tabular}
\end{table}
To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. 
Calcium cations at a molar concentration of $350\,\mathrm{mM}$ provide
$\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$ in the ECC-lipid model, which is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table SAMULI: They are now in the table \label{tab:binding} 
with contacts of calcium with different oxygens. However, I am not sure if there is enough information for the
table unless we analyze the contacts for Lipid14 models as well.}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which which predict a specific binding of sodium on 
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
which show the density profile of ions and water along the membrane normal. 
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}

@@

The contacts between 
%\ce{Ca^{2+}} 
ions and different oxygen moieties of POPC,
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by calculating the probability of 
having the smaller or equal distance than $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}.

@@

and \ce{Na+}
For sodium at $1000\,mM$ concentration 
the average number of contacts per lipid with any oxygen atom in POPC is 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
this number decreases to 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are counted 
it is even lower, 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
This means that interaction of sodium and POPC 
has balanced contributions of 55\% from only phosphate oxygens 
(i.e. no contacts with carbonyl oxyens),
and from only carbonyl oxygens with 20\% of configurations 
leaving 25\% for mixed phosphate and carbonyl oxygen configurations.  
This is in line with the calculated negative value of 
relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
suggesting that the populations of bound sodium cations 
naturally decrease with the decreasing density of water.
